RVD2: An ultra-sensitive variant detection model for low-depth targeted next-generation sequencing data

### 2013-10-03 doc/RECOMB2014/recomb2014.tex
create RECOMB2014 manuscript

\usepackage{setspace}
\doublespacing
%\usepackage{epstopdf}
%\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}

@@

\title[RVD2]{RVD2: An ultra-sensitive variant detection model for heterogeneous population sequencing}
\author{}
%\date{}                                           % Activate to display a given date or no date

@@

\section{Comparison with other variant calling algorithms}

GATK uses considerable preprocessing and a simple naive Bayes decision rule to call a variant. The genotype prior is fixed and constant across all loci and the likelihood of an allele at a locus is a function of the phred score \cite{McKenna2010}. Samtools

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=80mm]{pdf_figs/ROC_without_chi2.pdf}
\caption{ROC curve varying read depth showing detection performance of model with Bayesian Hypothesis Test.}
\label{fig:ROC}
\end{center}
\end{figure}

@@

\begin{figure}[h]
\begin{center}
\includegraphics[width=80mm]{pdf_figs/MuBarPlot.pdf}
\caption{Estimated minor allele fraction for called variants in 0.1\% dilution.}
\label{fig:MAF}
\end{center}
\end{figure}

\subsection{Comparison Results on Synthetic DNA}\

@@

\subsection{Pooled Multiple-Sclerosis Samples}\

>>>>

\usepackage{setspace}
%\doublespacing
%\usepackage{epstopdf}
%\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}

@@

\title[RVD2]{RVD2: An ultra-sensitive variant detection model for low-depth targeted next-generation sequencing data}
\author{}
%\date{}                                           % Activate to display a given date or no date

@@

\subsection{Comparison to other algorithms}

We compare the RVD2 model structure to other variant calling algorithms here. GATK uses considerable preprocessing and a simple naive Bayes decision rule to call a variant. The genotype prior is fixed and constant across all loci and the likelihood of an allele at a locus is a function of the phred score \cite{McKenna2010}.

[XXX Finish this text]

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=120mm]{pdf_figs/ROC_without_chi2.pdf}
\caption{ROC curve varying read depth showing detection performance of model with Bayesian Hypothesis Test.}
\label{fig:ROC}
\end{center}
\end{figure}

@@

\begin{figure}[h]
\begin{center}
\includegraphics[width=120mm]{pdf_figs/MuBarPlot.pdf}
\caption{Estimated minor allele fraction for called variants in 0.1\% dilution.}
\label{fig:MAF}
\end{center}
\end{figure}

\subsection{Empirical performance compared with other algorithms}

We compare the empirical performance of RVD2 to other variant calling algorithms. We use the synthetic DNA data sets described in the previous section to explore performance across a dynamic range of read depths as well as minor-allele frequencies. We report the sensitivity and specificity of each algorithm to detect the 14 true positive variant positions out of 400 tested positions.

GATK, samtools and VarScan2-mpileup are optimized to call genotypes on pure samples. Therefore, we expect those algorithms to perform well on the 100\% dilution (pure mutant) sample and poorly on heterogeneous samples. Varscan-somatic has been shown to outperform XXX,XXX, and XXX to detect a 5\% MAF for read depths between 100 and 1000. We show its performance on much lower MAF variants and across a wider range of coverage depths. Because Varscan2-somatic does not accept replicate data for the ``normal" or ``tumor" bam files, we used a single bam file from each group.

Table~\ref{t/l:comparison} shows the sensitivity and specificity for other variant calling algorithms. 
[XXX Finish this text]

@@

\subsection{Pooled Multiple-Sclerosis Samples}
